### 2014-07-25 paper.tex
updated text, rearranged figures and etc

% Please keep the abstract between 250 and 300 words
\section*{Abstract}
Marek's disease (MD) is caused by highly oncogenic Marek's disease virus (MDV).
Understanding the genetic basis for non-MHC resistance would be of both
fundamental and applied importance.  In this study, using computational
approaches of RNA-Seq data, we identified differentially expressed genes and
isoforms in response to MDV infection from highly inbred chickens lines that are
MD resistant and susceptible, and performed pathway and functional analysis.
Results show that expression of genes and isoforms involved in cell cytoskeleton
and adhesion molecules in line 6 may contribute to resistance of the disease by
limiting T cell activation, which in turn prevents the spread of MDV to
activated T cells.  In addition, we identified exonic single nucleotide
polymorphisms (SNPs) between lines 6 and 7 predicted their role in regulation of
alternative splicing.  Based on our results, we speculate that alternative
splice forms of cell adhesion molecules and cytoskeleton play and important role
in controlling T cell activation in line 6.

@@

Marek's disease (MD) is an economically significant chicken disease that affects
the poultry industry worldwide with estimated annual loss of \$2 billion loss
has been reported from outbreaks~\cite{morrow2004marek}.  The disease is caused
by the highly oncogenic Marek's disease virus (MDV), an alphaherpesvirus that
induces T-cell lymphomas in susceptible birds.  Vaccination is the primary
control measure, which is effective in reducing incidence of tumor formation.
However, since MD vaccines are not sterilizing, they do not prevent infection or
horizontal spread of the virus.  As a consequence, MDV field strains that
overcome vaccinal protection have arisen repeatedly over time.  Therefore, there
is a need for sustainable alternative controls measures, such as improving
genetic resistance.

Many studies have reported strong associations between MHC alleles and
resistance or susceptibility to MD.  For example, chickens with MHC allele
B\textsuperscript{21} are highly resistant in contrast to chickens with
B\textsuperscript{19} allele which are highly susceptible.  ADOL lines 6 and 7,
both share (B\textsuperscript{2}) allele, yet exhibit different phenotypic
responses to MD; e.g., challenge with JM/102W strain with result in 0 and 100\%
MD incidence for lines 6 and 7, respectively.  Thus, the major unanswered
questions are what genetic factors, especially those that are non-MHC,
contribute to susceptibility and resistance of the disease and what is the main
contributing mechanism.

In the past decades, significant efforts have been spent to study variations in
a global gene expression between resistant and susceptible birds using
microarray and RNA-Seq methods in order to identify non-MHC genes that
contribute to resistance to
MD~\cite{sarson2008transcriptional,morgan2001induction,vallejo1998genetic,yonash1999high,bumstead1998genomic}.
However, none of the studies have investigated differential expression of
alternative isoforms, which are known to play a significant role in many
biological events including immune responses.  In addition, studies have shown
that isoform expression levels can provide better signatures for some
diseases~\cite{zhang2013isoform}.  Changes of isoform expression levels are
governed partly by two types of {\em cis-}regulatory elements: Exon Splicing
Enhancer (ESE) and Exon Splicing Silencer (ESS), which are located within an
exon sequence.  A number of sequence motifs of ESE and ESS have been identified
in human and some other organisms and could be predicted {\em in silico}.
Mutations that disrupt or create those motifs could alter splicing patterns
leading to aberrant alternative splicing.  A number of disease-associated
single-nucleotide polymorphsims in coding regions (SNPs) that affect ESEs and
ESSs have been well characterized~\cite{blencowe2000exonic, wang2007splicing}.
%In fact, 15\% of mutations that cause genetic diseases affect pre-mRNA splicing
%via this mechanism.
Therefore, variations in isoform expression could lead to identification of SNPs
that underlie genetic resistance to MD.  In this article, we reported
differential-expressed genes and isoforms that may contribute to resistance to
MD as well as SNPs that can potentially affect isoform expression levels.

@@

Some DE genes have been reported from previous microarray studies were also
found to be differentially expressed in this study.  For example, B6.1 (Bu-1)
gene is known to be down-regulated approximately 2.27 fold in susceptible
chicken with the MHC allele B\textsuperscript{19} at 4
d.p.i~\cite{sarson2008transcriptional}.  It was also found to be down-regulated
about 3-fold in line 7 (susceptible).  Similarly, {\em GMZA} gene reported to be
upregulated across genetically different chickens (B\textsuperscript{19},
B\textsuperscript{21} alleles),
was also found to be highly upregulated.  In contrast, some genes that have been
reported to be highly expressed in resistant chickens were downregulated in both
lines. Those genes are {\em AMIGO2, MMP13} and {\em CLEC3B}, which were found to
be downregulated more than 2-fold~\cite{sarson2008transcriptional}.
% Several genes involved in innate immune response were differentially expressed
% in resistant chickens: \textit{DDT, NMU, VIPR2} and \textit{HSP5}. Of all
% those genes, only HSP5 was upregulated.
Other immune genes reported to be highly expressed in susceptible chickens (line
7) including {\em AVD, ART1, NOS2, CXCL13L2, MX1} and {\em
SOCS1}~\cite{smith2011systems} were also found to be highly upregulated in both
line 6 and 7.  However, our results show the similar expression patterns for
{\em IL6} and {\em IL18}, which were only upregulated in line 7 in early stage
of infection ($3-5$ d.p.i).

In contrast, {\em IL15} has been reported non differential-expressed between
control and infected chickens in line 6 and 7~\cite{kaiser2003differential};
however, it was only upregulated in line 7 in our study.  Expression of {\em
IL15} is induced by {\em TLR9}, which binds to non-methylated CpG residues
present in genomes of many DNA viruses, including herpes simplex virus.  This
cytokine auto-regulates the expression of {\em CD40}, which is a transmembrane
receptor required for activation of macrophages by CD4 T cells.  Consequently,
{\em CD40} was only upregulated in line 7 (data not shown).

\subsection*{Differential gene expression indicates active immune responses to
ongoing lytic infection in line 7}

Many genes were found to be differentially expressed (DE) between control and
infected chickens in lines 6 and line 7.  While the number of unique
downregulated genes in both lines was approximately equal, the number of unique
upregulated genes in line 7 was much greater compared to line 6
(Figure~\ref{degenes_venn}).  Interestingly, some genes
that were differentially expressed in both lines were regulated in the opposite
direction (Table~\ref{tab:opposite}).  Among genes downregulated in line 6 but
upregulated in line 7 were {\em LL} (lung lectin) and {\em SFTPA1}, which encode
a calcium-dependent C-type lectin and a lung surfactant protein respectively.
Both molecules are important in innate
immunity~\cite{hogenkamp2008chicken,kingma2006defense}.  {\em LIMS1} is involved
in cell differentiation and proliferation and {\em PPARG} is a suppressor of
{\em NF$\kappa$B}-mediated proinflamatory response.  On the other hand, nearly
all genes upregulated in line 6 but downregulated in line 7 are involved in cell
survival such as mRNAs splicing, cell growth, and protein synthesis, except CD7
whose function is involved in T cell-B cell interaction.  This difference
suggests that even at this stage of infection in line 6, the lytic phase could
be subsided, therefore, only genes important for survival of cell are
upregulated in line 6.  In comparison, the lytic phase in line 7 may still
continue and as a result, genes involved in immune responses are still
upregulated.  In addition, type I interferon ({\em IFN-$\gamma$} and {\em
IFN-$\beta$}) as well as {\em INF-$\alpha$3} were found to be highly upregulated
in infected chickens in both lines 6 and 7 (Table~\ref{tab:cytokines}).
However, expressions of genes encoding their corresponding receptors were not
different in line 6, but upregulated in line 7.  This could also reflect the
ongoing immune responses in line 7.

\subsection*{Functional analysis of differential-expressed genes indicates
inactive adaptive immune responses in line 6}

To determine pathways that were perturbed during the infection, data were
analyzed by GOSeq, which accounted for gene lengths bias unique for RNA-Seq
data~\cite{young2010method}.  Significantly perturbed pathways (FDR $< 0.1$)
from both lines that involved in immune response include TLR signaling pathway,
cytokine-cytokine receptor interaction, intestinal immune network for IgA
production, and cell adhesion molecules (CAMs)(figure~\ref{line67_kegg}).  Some
other pathways important in response to viral infection and only significantly
enriched in line 7 include phagosome, apoptosis, RIG-I-like receptor signaling
pathway, NOD-like receptor signaling pathway, and lysosome.
Figure~\ref{kegg_phagosome} is a pictorial example of differentially expressed
genes in the phagosome pathway.  Although MHC I gene ({\em BF1}) was
differentially expressed in both lines, other genes involved in expressing newly
synthesized MHC I were only upregulated in line 7 suggesting that new MHC I
molecules were actively produced.  Furthermore, gene ontology analysis of
biological processes (GO:BP) shows that categories involved in both adaptive and
innate immune responses were enriched in line 7 (supplementary material).  On
the other hand, only categories involved in innate immune responses were
enriched in line 6.  In addition, enrichment of apoptosis pathway in line 7
indicates that the programmed cell death could be induced by CTL response to
eliminate the viruses.

At this stage of infection, our results suggest that lytic infection of MDV
stimulates both innate and adaptive immune responses, which leads to activation
of T cells in line 7.  Only activated T cells are infected by MDV, therefore,
the lytic phase could facilitate the spread of the viruses by enhancing
expansion of activated T cells.  Due to cell-associate nature of MDV, the
viruses transfer to T cells via cell-to-cell contact between B cells and T cells
during antigen presentation or B cells activation by T helper cells.  Therefore,
it is beneficial for the host to restrain such contact.  However, it is not
clear how chickens in line 6 control the lytic infection of MDV.  Two mechanisms
have been speculated to contribute to MD resistance.  First innate immune
responses could be highly effective and could activate strong adaptive immune
responses that rapidly control viral replication and force the viruses to
undergo latent phase.  Second, innate immune resposnes itself could be highly
effective in limiting viral replication~\cite{smith2011systems}.

@@

The immune system is isoform-rich and many genes express different isoforms with
distinctive functions in response to stimuli such as stress, chemicals and
infection.  Changes in expression of splice forms of immune related genes have
been reported to be associated with increase susceptibility and poor prognosis
of diseases~\cite{lynch2004consequences}.  Studying differential isoform
expression could therefore shed lights into inherent differences between lines 6
and 7 that confer resistance or susceptibility to MD.

@@

Due to those issues, it is not feasible to accurately estimate expression of
isoforms, especially when gene annotation is constructed from {\em de novo}
assembly~\cite{trapnell2013differential}.  To avoid these issues, we chose to
study exon expression instead of isoform expression.  Using MISO with
exon-centric method, only reads spanning across a few exons are used and only
exons involved in a splicing event are examined.  The expression of exon
inclusion is calculated as Percent Spliced In (Psi or $\Psi$), which can be used
to infer the portion of transcripts that include the exon in each
sample~\cite{Katz:2010iv}.  In this study, we investigated the three most common
alternative splicing events in vertebrates, which are skipped exons (SE), an
alternative $3\prime$ (A3SS) and $5\prime$ (A5SS) splice site.
Lists of DEU genes from line 6 that show difference in $\Psi$ greater than 0.20
when compared to line 7 in infected chickens are shown in
Tables~\ref{tab:line67i_diff_line67u_one},~\ref{tab:line67i_diff_line67u_two}
and~\ref{tab:line67i_diff_line67u_three}.  Genes can be categorized roughly into
four groups based on the pattern of $\Psi$ across control and infected birds in
both lines.

Group I (Table~\ref{tab:line67i_diff_line67u_one}) includes genes with $\Psi$s
that were up- or down-regulated in infected chickens in line 6 only.  This group
includes {\em BCL11B} (B-cell CLL/lymphoma 11B zinc finger proteins), a B-cell
lymphoma associated C2H2-type zinc finger protein encoding gene, which functions
as a tumour-suppressor in T-cell lymphoma in human.  According to homologous
alignments on UCSC genome browser, a splice form with the skipped exon is
similar to mouse {\em BCL11B isoform b}.  The skipped exon was expressed 30\% in
infected line 6 chickens; whereas it was rarely expressed (4-7\%) in the control
line 6 and both groups in line 7.  The skipped exon was not found to encode any
known protein domain, however, it is in the middle of two adjacent C2H2-type
finger protein domains.  Another gene related to B and T cell lymphomas is {\em
SIK2} (salt-inducible kinase 2).  This gene has been reported to have a negative
effect on T cell lymphomas by limiting the transcription of HTLV-1
virus~\cite{tang2013lkb1}.  Some other genes are important in pre-mRNA splicing
including {\em TRA2A}, {\em SRSF6} and {\em GEMIN6}.  {\em TRA2A} (transformer-2
protein homolog alpha) encodes RNA recognition motif (RRM), the skipped exon is
not found to encode any known protein domain.  {\em SRSF6} (SR splicing factor
6) encodes a nuclear protein that belongs to the splicing factor protein family.
{\em GEMIN6} plays a role in the assembly of spliceosomal snRNP in cytoplasm.
These genes could play a significant role in regulating inclusion of alternative
exons.  A gene in this group that might be important for innate immune responses
is {\em RAC3} (Ras-related C3 botulinum toxin substrate 3).  This gene encodes
small GTPases, belonging to Ras family, that regulates a wide variety of
cellular events including cell growth, cytoskeletal reorganization, and the
activation of protein kinases.  The role of small GTPases in immune responses is
discussed further below.

@@

Group III (Table~\ref{tab:line67i_diff_line67u_three}) includes genes that
exhibit differential isoform expression only in infected line 7.  A number of
genes in this group encode proteins that are parts of spliceosome: {\em SRSF3},
{\em HNRNPDL}, {\em SFSWAP}, {\em THOC1}, {\em RNPC3} and {\em SRSF5}.  Two
genes involved in cell-cell contact regulated by integrins are also in this
group.  {\em PPP1R12A} is a myosin phosphatase that regulates the interaction of
actin and myosin downstream of GTPase Rho proteins.  {\em PODXL} encodes
PODX-like protein that functions in integrin-dependent manner as both
pro-adhesive and anti-adhesive molecules.  This protein is involved in
cell-to-cell contact, cell trafficking, and cancer
progression~\cite{nielsen2009role,somasiri2004overexpression}.

The last group (Group IV, Table~\ref{tab:line67i_diff_line67u_three}) only has
one gene, {\em GOSR1}.  The $\Psi$ value of this gene was less than 0.20 cutoff
in control and infected chickens in line 6, but it is significantly different
between infected chickens in line 6 and 7.  Also, there is a significant
difference between control and infected chickens in line 7.  This gene encodes a
trafficking membrane protein important for transporting proteins from {\em cis-}
to {\em trans-}golgi network.

@@

Absence of LFA-1 leads to impaired functions of lymphocytes in proliferation and
tumor rejection~\cite{scharffetter1998spontaneous,schmits1996lfa}.  Mutations in
{\em ITGB2} gene has been associated with type 1 leukocyte adhesion deficiency
(LAD-1), an autosomal-recessive inherited disease found in few families.  The
disease is characterized by impair of lymphocytes in adherent-dependent
functions, lack of accumulation to the site of infection and recurrent bacterial
and fungal infection~\cite{springer1987lymphocyte}.  In addition, the response
of lymphocytes to mitogens is decreased in patients with
LAD~\cite{springer1987lymphocyte}.  The decrease in responsiveness to mitogens
have been shown to correlate with resistance to MD by Lee and
Bacon~\cite{lee1983ontogeny}, who illustrated that resistant birds (line 6 and
N) were less responsive to phytohemagglutinin (PHA) than susceptible birds (line
7 and P).

Actin cytoskeleton is very important in T cell activation because it enhances
activity of LFA-1 by increasing its avidity and recruiting signaling molecules
necessary for downstream signaling~\cite{dustin2000immunological,
van2000avidity}.  Cytoskeleton proteins binding to cytoplasmic domain of LFA-1
are thought to play an important role in driving LFA-1 to aggregation on the
cell surface, resulting in increased avidity.  Aggregation of LFA-1 has been
demonstrated to be essential for lymphocytes to bind to the
ligand~\cite{van1994extracellular}.  Interestingly, {\em RAC3, PFN2,} and {\em
PPP1R12A}, which are involved in actin cytoskeleton pathway
(Figure~\ref{kegg_actin}), also expressed different ratios of alternative splice
forms between lines 6 and 7.  Some of these genes are also co-present in three
other pathways that involved in immune responses (Table~\ref{tab:integrin}).
It could be speculated that pre-mRNA splicing of these genes is co-regulated
by splicing regulators or some genetic factors.

@@

Even though the exact mechanism is not known, incompetency of T cells in
response to stimuli appears to benefit resistant birds because MD virus could
not induce T cells to proliferate and cause them to undergo neoplastic
transformation.  It has also been suggested that the mechanism that controls
both lymphocyte proliferation induced by MD virus and lymphocyte proliferation
induced by immune response is the same~\cite{pazderka1975histocompatibility}.
Therefore, it may be useful to consider a link between deficiency of lymphocytes
in line 6 to the alternative splice form of {\em ITGB2} that is only expressed
in line 6.  Although the exon included in the alternative splice form is
non-coding, it could serve important functions in translation or
posttranscriptional regulation.

% Based on our gene model, the skipped exon of \textit{ITGB2} is located in the
% 5$\prime$ UTR; whereas, the skipped exon of \textit{HCK}, encodes protein
% tyrosine kinase (Pfam:PF07714).

@@

Although alternative splicing of genes in all groups could be regulated by {\em
cis-} and {\em trans-}splicing factors, genes in group II are more likely to be
regulated by {\em cis-}regulatory factors.  The ratios of isoform expression in
this group were relatively stable within line, but were significantly different
between lines.  Investigation of nucleotide differences within exons of lines 6
and 7 could reveal a possible role of SNPs in regulating alternative splicing in
this group.  We obtained a sequence of an alternative exon from line 6 and used
Human Splicing Finder (HSF) to determine whether SNPs from line 7 could alter
predicted ESEs or ESSs.  Results from some genes involved in cytokinesis are
discussed in this section.  Exonic SNPs from line 6 and 7 are listed in
Table~\ref{tab:deu_snps}.

>>>>

% Please keep the abstract between 250 and 300 words
\section*{Abstract}
Marek's disease (MD) is caused by highly oncogenic Marek's
disease virus (MDV).  Understanding the genetic basis for non-MHC
resistance would be of both fundamental and applied importance.
In this study, using computational approaches of RNA-Seq data, we
identified differentially expressed genes and isoforms in
response to MDV infection from highly inbred chickens lines that
are MD resistant (line 6), susceptible (line 7), and performed
pathway and functional analysis.  Results show that expression of
genes and isoforms involved in cell cytoskeleton and adhesion
molecules in the resistant line may contribute to resistance of
the disease by limiting T cell activation, which in turn prevents
the spread of MDV to activated T cells.  In addition, we
identified exonic single nucleotide polymorphisms (SNPs) between
lines predicted role in regulation of alternative splicing.
Based on our results, we speculate that alternative splice forms
of cell adhesion molecules and cytoskeleton play and important
role in controlling T cell activation in the resistant line.

@@

Marek's disease (MD) is an economically significant chicken
disease that affects the poultry industry worldwide with
estimated annual of \$2 billion~\cite{morrow2004marek}.  The
disease is caused by the highly oncogenic Marek's disease virus
(MDV), an alphaherpesvirus that induces T-cell lymphomas in
susceptible birds.  Vaccination is the primary control measure,
which is effective in reducing incidence of tumor formation.
However, since MD vaccines are not sterilizing, they do not
prevent infection or horizontal spread of the virus.  As a
consequence, MDV field strains that overcome vaccinal protection
have arisen repeatedly over time.  Therefore, there is a need for
sustainable alternative controls measures, such as improving
genetic resistance.

Many studies have reported strong associations between MHC
alleles and resistance or susceptibility to MD.  For example,
chickens with MHC allele B\textsuperscript{21} are highly
resistant in contrast to chickens with B\textsuperscript{19}
allele which are highly susceptible.  ADOL lines 6 and 7, both
share (B\textsuperscript{2}) allele, yet exhibit different
phenotypic responses; e.g., challenge with JM/102W strain result
in 0 and 100\% MD incidence for lines 6 and 7, respectively.
Thus, the major unanswered questions are what genetic factors,
especially those that are non-MHC, contribute to susceptibility
and resistance of the disease and what is the main contributing
mechanism.

In the past decades, significant efforts have been to study
variations in a global gene expression between resistant and
susceptible birds using microarray and RNA-Seq methods in order
to identify non-MHC genes that contribute to resistance to
MD~\cite{sarson2008transcriptional,morgan2001induction,vallejo1998genetic,yonash1999high,bumstead1998genomic}.
However, none of the studies have investigated differential
expression of alternative isoforms, which are known to play a
significant role in many biological events including immune
responses.  In addition, studies have shown that isoform
expression levels can provide better signatures for some
diseases~\cite{zhang2013isoform}.  Changes of isoform expression
levels are governed partly by two types of {\em cis-}regulatory
elements: Exon Splicing Enhancer (ESE) and Exon Splicing Silencer
(ESS), which are located within an exon sequence.  A number of
sequence motifs of ESE and ESS have been identified in human and
some other organisms and could be predicted {\em in silico}.
Mutations that disrupt or create those motifs could alter
splicing patterns leading to aberrant alternative splicing.  A
number of disease-associated single-nucleotide polymorphsims in
coding regions (SNPs) that affect ESEs and ESSs have been well
characterized~\cite{blencowe2000exonic, wang2007splicing}.
%In fact, 15\% of mutations that cause genetic diseases affect
%pre-mRNA splicing via this mechanism.
Therefore, variations in isoform expression could lead to
identification of SNPs that underlie genetic resistance to MD.
In this article, we reported differential-expressed genes and
isoforms that may contribute to resistance to MD as well as SNPs
that can potentially affect isoform expression levels.

@@

Some DE genes have been reported from previous microarray studies
were also found to be differentially expressed in this study.
For example, B6.1 (Bu-1) is known to be down-regulated
approximately 2.27 fold in susceptible chickens with the MHC
allele B\textsuperscript{19} at 4 days post infection
(d.p.i)~\cite{sarson2008transcriptional}.  It was also found to
be down-regulated about 3-fold in the susceptible line in our
study.  Similarly, {\em GMZA} reported to be upregulated across
genetically different chickens (B\textsuperscript{19},
B\textsuperscript{21} alleles), was also found to be highly
upregulated.  In contrast, some genes that have been reported to
be highly expressed in resistant chickens were downregulated in
both lines. Those genes are {\em AMIGO2, MMP13} and {\em CLEC3B},
which were found to be downregulated more than
2-fold~\cite{sarson2008transcriptional}.
% Several genes involved in innate immune response were
% differentially expressed in resistant chickens: \textit{DDT,
% NMU, VIPR2} and \textit{HSP5}. Of all those genes, only HSP5
% was upregulated.
Other immune genes reported to be highly expressed in susceptible
chickens including {\em AVD, ART1, NOS2, CXCL13L2, MX1} and {\em
SOCS1}~\cite{smith2011systems} were also found to be highly
upregulated in both lines.  However, our results show the similar
expression patterns for {\em IL6} and {\em IL18}, which were only
upregulated in the susceptible line in early stage of infection
($3-5$ d.p.i).

In contrast, {\em IL15} has been reported non
differential-expressed between control and infected chickens in
both lines~\cite{kaiser2003differential}; however, it was only
upregulated in the susceptible line.  Expression of {\em IL15} is
induced by {\em TLR9}, which binds to non-methylated CpG residues
present in genomes of many DNA viruses, including herpes simplex
virus.  This cytokine auto-regulates the expression of {\em
CD40}, which is a transmembrane receptor required for activation
of macrophages by CD4 T cells.  Consequently, {\em CD40} was only
upregulated in the susceptible line (data not shown).

\subsection*{Differential gene expression indicates active immune
responses to ongoing lytic infection in the susceptible line}

Many genes were found to be differentially expressed (DE) between
control and infected chickens in both lines.  While the number of
unique downregulated genes in both lines was approximately equal,
the number of unique upregulated genes in the susceptible line
was much greater compared to the resistant line
(Figure~\ref{degenes_venn}).  Interestingly, some genes that were
differentially expressed in both lines were regulated in the
opposite direction (Table~\ref{tab:opposite}).  Among genes
downregulated in the resistant line but upregulated in the
susceptible line were {\em LL} (lung lectin) and {\em SFTPA1},
which encode a calcium-dependent C-type lectin and a lung
surfactant protein respectively.  Both molecules are important in
innate immunity~\cite{hogenkamp2008chicken,kingma2006defense}.
{\em LIMS1} is involved in cell differentiation and proliferation
and {\em PPARG} is a suppressor of {\em NF$\kappa$B}-mediated
proinflamatory response.  On the other hand, nearly all genes
upregulated in the resistant line but downregulated in the
susceptible line are involved in cell survival such as mRNAs
splicing, cell growth, and protein synthesis, except CD7 whose
function is involved in T cell-B cell interaction.  This
difference suggests that even at this stage of infection in the
resistant line, the lytic phase could be subsided, therefore,
only genes important for survival of cell are upregulated in the
resistant line.  In comparison, the lytic phase in the
susceptible line may still continue and as a result, genes
involved in immune responses are still upregulated.  In addition,
type I interferon ({\em IFN-$\gamma$} and {\em IFN-$\beta$}) as
well as {\em INF-$\alpha$3} were found to be highly upregulated
in infected chickens in both lines  (Table~\ref{tab:cytokines}).
However, expressions of genes encoding their corresponding
receptors were not different in the resistant line, but
upregulated in the susceptible line.  This could also reflect the
ongoing immune responses in the susceptible line.

\subsection*{Functional analysis of differential-expressed genes
indicates inactive adaptive immune responses in the resistant
line}

To determine pathways that were perturbed during the infection,
data were analyzed by GOSeq, which accounted for gene lengths
bias unique for RNA-Seq data~\cite{young2010method}.
Significantly perturbed pathways (FDR $< 0.1$) from both lines
that involved in immune response include TLR signaling pathway,
cytokine-cytokine receptor interaction, intestinal immune network
for IgA production, and cell adhesion molecules
(CAMs)(figure~\ref{line67_kegg}).  Some other pathways important
in response to viral infection and only significantly enriched in
the susceptible line include phagosome, apoptosis, RIG-I-like
receptor signaling pathway, NOD-like receptor signaling pathway,
and lysosome.  Figure~\ref{kegg_phagosome} is a pictorial example
of genes in the phagosome pathway.  Although MHC I ({\em BF1})
was differentially expressed in both lines, other genes involved
in expressing newly synthesized MHC I were only upregulated in
the susceptible line suggesting that new MHC I molecules were
actively produced.  Furthermore, gene ontology analysis of
biological processes (GO:BP) shows that categories involved in
both adaptive and innate immune responses were enriched in the
susceptible line (supplementary material).  On the other hand,
only categories involved in innate immune responses were enriched
in the resistant line.  In addition, enrichment of apoptosis
pathway in the susceptible line indicates that the programmed
cell death could be induced by CTL response to eliminate the
viruses.

At this stage of infection, our results suggest that lytic
infection of MDV stimulates both innate and adaptive immune
responses, which leads to activation of T cells in the
susceptible line.  Only activated T cells are believed to be
infected by MDV, therefore, the lytic phase could facilitate the
spread of the viruses by enhancing expansion of activated T
cells.  Due to cell-associate nature of MDV, the viruses transfer
to T cells via cell-to-cell contact between B cells and T cells
during antigen presentation or B cells activation by T helper
cells.  Therefore, it is beneficial for the host to restrain such
contact.  However, it is not clear how chickens in the resistant
line control the lytic infection of MDV.  Two mechanisms have
been speculated to contribute to MD resistance.  First innate
immune responses could be highly effective and could activate
strong adaptive immune responses that rapidly control viral
replication and force the viruses to undergo latent phase.
Second, innate immune resposnes itself could be highly effective
in limiting viral replication~\cite{smith2011systems}.

@@

The immune system is isoform-rich and many genes express
different isoforms with distinctive functions in response to
stimuli such as stress, chemicals and infection.  Changes in
expression of splice forms of immune related genes have been
reported to be associated with increase susceptibility and poor
prognosis of diseases~\cite{lynch2004consequences}.  Studying
differential isoform expression could therefore shed lights into
inherent differences between lines that confer resistance or
susceptibility to MD.

@@

Due to those issues, it is not feasible to accurately estimate
expression of isoforms, especially when gene annotation is
constructed from {\em de novo}
assembly~\cite{trapnell2013differential}.  To avoid these issues,
we chose to study exon expression instead of isoform expression.
Using MISO with exon-centric method, only reads spanning across a
few exons are used and only exons involved in a splicing event
are examined.  The expression of exon inclusion is calculated as
Percent Spliced In (Psi or $\Psi$), which can be used to infer
the portion of transcripts that include the exon in each
sample~\cite{Katz:2010iv}.  In this study, we investigated the
three most common alternative splicing events in vertebrates,
which are skipped exons (SE), an alternative $3\prime$ (A3SS) and
$5\prime$ (A5SS) splice site.  Lists of DEU genes from the
resistant line that show difference in $\Psi$ greater than 0.20
when compared to the susceptible line in infected chickens are
shown in
Tables~\ref{tab:line67i_diff_line67u_one},~\ref{tab:line67i_diff_line67u_two}
and~\ref{tab:line67i_diff_line67u_three}.  Genes can be
categorized roughly into four groups based on the pattern of
$\Psi$ across control and infected birds in both lines.

Group I (Table~\ref{tab:line67i_diff_line67u_one}) includes genes
with $\Psi$s that were up- or down-regulated in infected chickens
in the resistant line only.  This group includes {\em BCL11B}
(B-cell CLL/lymphoma 11B zinc finger proteins), a B-cell lymphoma
associated C2H2-type zinc finger protein encoding gene, which
functions as a tumour-suppressor in T-cell lymphoma in human.
According to homologous alignments on UCSC genome browser, a
splice form with the skipped exon is similar to mouse {\em BCL11B
isoform b}.  The skipped exon was expressed 30\% in infected the
resistant line chickens; whereas it was rarely expressed (4-7\%)
in the control the resistant line and both groups in the
susceptible line.  The skipped exon was not found to encode any
known protein domain, however, it is in the middle of two
adjacent C2H2-type finger protein domains.  Another gene related
to B and T cell lymphomas is {\em SIK2} (salt-inducible kinase
2).  This gene has been reported to have a negative effect on T
cell lymphomas by limiting the transcription of HTLV-1
virus~\cite{tang2013lkb1}.  Some other genes are important in
pre-mRNA splicing including {\em TRA2A}, {\em SRSF6} and {\em
GEMIN6}.  {\em TRA2A} (transformer-2 protein homolog alpha)
encodes RNA recognition motif (RRM), the skipped exon is not
found to encode any known protein domain.  {\em SRSF6} (SR
splicing factor 6) encodes a nuclear protein that belongs to the
splicing factor protein family.  {\em GEMIN6} plays a role in the
assembly of spliceosomal snRNP in cytoplasm.  These genes could
play a significant role in regulating inclusion of alternative
exons.  A gene in this group that might be important for innate
immune responses is {\em RAC3} (Ras-related C3 botulinum toxin
substrate 3).  This gene encodes small GTPases, belonging to Ras
family, that regulates a wide variety of cellular events
including cell growth, cytoskeletal reorganization, and the
activation of protein kinases.  The role of small GTPases in
immune responses is discussed further below.

@@

Group III (Table~\ref{tab:line67i_diff_line67u_three}) includes
genes that exhibit differential isoform expression only in
infected the susceptible line.  A number of genes in this group
encode proteins that are parts of spliceosome: {\em SRSF3}, {\em
HNRNPDL}, {\em SFSWAP}, {\em THOC1}, {\em RNPC3} and {\em SRSF5}.
Two genes involved in cell-cell contact regulated by integrins
are also in this group.  {\em PPP1R12A} is a myosin phosphatase
that regulates the interaction of actin and myosin downstream of
GTPase Rho proteins.  {\em PODXL} encodes PODX-like protein that
functions in integrin-dependent manner as both pro-adhesive and
anti-adhesive molecules.  This protein is involved in
cell-to-cell contact, cell trafficking, and cancer
progression~\cite{nielsen2009role,somasiri2004overexpression}.

The last group (Group IV,
Table~\ref{tab:line67i_diff_line67u_three}) only has one gene,
{\em GOSR1}.  The $\Psi$ value of this gene was less than 0.20
cutoff in control and infected chickens in the resistant line,
but it is significantly different between infected chickens in
the resistant and susceptible lines.  Also, there is a significant
difference between control and infected chickens in the
susceptible line.  This gene encodes a trafficking membrane
protein important for transporting proteins from {\em cis-} to
{\em trans-}golgi network.

@@

Absence of LFA-1 leads to impaired functions of lymphocytes in
proliferation and tumor
rejection~\cite{scharffetter1998spontaneous,schmits1996lfa}.
Mutations in {\em ITGB2} gene has been associated with type 1
leukocyte adhesion deficiency (LAD-1), an autosomal-recessive
inherited disease found in few families.  The disease is
characterized by impair of lymphocytes in adherent-dependent
functions, lack of accumulation to the site of infection and
recurrent bacterial and fungal
infection~\cite{springer1987lymphocyte}.  In addition, the
response of lymphocytes to mitogens is decreased in patients with
LAD~\cite{springer1987lymphocyte}.  The decrease in
responsiveness to mitogens have been shown to correlate with
resistance to MD by Lee and Bacon~\cite{lee1983ontogeny}, who
illustrated that resistant birds (the resistant line and N) were
less responsive to phytohemagglutinin (PHA) than susceptible
birds (line 7 and P).

Actin cytoskeleton is very important in T cell activation because
it enhances activity of LFA-1 by increasing its avidity and
recruiting signaling molecules necessary for downstream
signaling~\cite{dustin2000immunological, van2000avidity}.
Cytoskeleton proteins binding to cytoplasmic domain of LFA-1 are
thought to play an important role in driving LFA-1 to aggregation
on the cell surface, resulting in increased avidity.  Aggregation
of LFA-1 has been demonstrated to be essential for lymphocytes to
bind to the ligand~\cite{van1994extracellular}.  Interestingly,
{\em RAC3, PFN2,} and {\em PPP1R12A}, which are involved in actin
cytoskeleton pathway (Figure~\ref{kegg_actin}), also expressed
different ratios of alternative splice forms between lines .
Some of these genes are also co-present in three other pathways
that involved in immune responses (Table~\ref{tab:integrin}).  It
could be speculated that pre-mRNA splicing of these genes is
co-regulated by splicing regulators or some genetic factors.

@@

Even though the exact mechanism is not known, incompetency of T
cells in response to stimuli appears to benefit resistant birds
because MDV could not induce T cells to proliferate and cause
them to undergo neoplastic transformation.  It has also been
suggested that the mechanism that controls both lymphocyte
proliferation induced by MDV and lymphocyte proliferation induced
by immune response is the
same~\cite{pazderka1975histocompatibility}.  Therefore, it may be
useful to consider a link between deficiency of lymphocytes in
the resistant line to the alternative splice form of {\em ITGB2}
that is only expressed in the resistant line.  Although the exon
included in the alternative splice form is non-coding, it could
serve important functions in translation or posttranscriptional
regulation.

% Based on our gene model, the skipped exon of \textit{ITGB2} is
% located in the 5$\prime$ UTR; whereas, the skipped exon of
% \textit{HCK}, encodes protein tyrosine kinase (Pfam:PF07714).

@@

Although alternative splicing of genes in all groups could be
regulated by {\em cis-} and {\em trans-}splicing factors, genes
in group II are more likely to be regulated by {\em
cis-}regulatory factors.  The ratios of isoform expression in
this group were relatively stable within line, but were
significantly different between lines.  Investigation of
nucleotide differences within exons of both lines could reveal a
possible role of SNPs in regulating alternative splicing in this
group.  We obtained a sequence of an alternative exon from the
resistant line and used Human Splicing Finder (HSF) to determine
whether SNPs from the susceptible line could alter predicted ESEs
or ESSs.  Results from some genes involved in cytokinesis are
discussed in this section.  Exonic SNPs from the resistant and
susceptible lines are listed in Table~\ref{tab:deu_snps}.
